D3064R0: OOTA Execution is Provably Vacuous

### 2023-12-02 ootavacuous.tex
Add the first five causality tests
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

Although non-volatile accesses to atomic objects are not observable behavior,
they are still subject to the memory model.
In particular, implementations must not behave as if they have:

@@

% \appendix

% \section{Demonstration of OOTA-Cycle Definition}
% \label{app:Demonstration of OOTA-Cycle Definition}

>>>>

Although non-volatile accesses to atomic objects are not observable behavior,
they are still subject to the memory model.
In particular, implementations must not behave as if the source code has
additional:

@@

\appendix

\section{Demonstration of OOTA-Cycle Definition}
\label{app:Demonstration of OOTA-Cycle Definition}

\subsection{Litmus Tests from “Causality Test Cases"}
\label{app:Litmus Tests from “Causality Test Cases"}

\subsubsection{Causality Test Case 1}
\label{app:Causality Test Case 1}

Listing~\ref{lst:Causality Test Case 1}
shows causality test case 1, for which the \co{r1 == r2 == 1} result
is to be allowed.
However, the herd7 tool does not find this litmus test's OOTA cycle.

\begin{listing}[tbp]
\begin{verbatim}
 1 C oota-causality-1
 2 {
 3   [x] = 0;
 4   [y] = 0;
 5 }
 6
 7 P0(volatile atomic_int *x, volatile atomic_int *y) {
 8   int r1 = atomic_load_explicit(x, memory_order_relaxed);
 9   if (r1 >= 0)
10     atomic_store_explicit(y, 1, memory_order_relaxed);
11 }
12
13 P1(volatile atomic_int *x, volatile atomic_int *y) {
14   int r2 = atomic_load_explicit(y, memory_order_relaxed);
15   atomic_store_explicit(x, r1, memory_order_relaxed);
16 }
17
18 exists(0:r1=1 /\ 1:r2=1)
\end{verbatim}
\caption{Causality Test Case 1}
\label{lst:Causality Test Case 1}
\end{listing}

\subsubsection{Causality Test Case 2}
\label{app:Causality Test Case 2}

Listing~\ref{lst:Causality Test Case 2}
shows causality test case 2, for which the \co{r1 == r2 == r3 == 1} result
is to be allowed.
The herd7 tool finds this OOTA cycle.

\begin{listing}[tbp]
\begin{verbatim}
 1 C oota-causality-2
 2 {
 3   [x] = 0;
 4   [y] = 0;
 5 }
 6
 7 P0(volatile atomic_int *x, volatile atomic_int *y) {
 8   int r1 = atomic_load_explicit(x, memory_order_relaxed);
 9   int r2 = atomic_load_explicit(x, memory_order_relaxed);
10   if (r1 == r2)
11     atomic_store_explicit(y, 1, memory_order_relaxed);
12 }
13
14 P1(volatile atomic_int *x, volatile atomic_int *y) {
15   int r3 = atomic_load_explicit(y, memory_order_relaxed);
16   atomic_store_explicit(x, r3, memory_order_relaxed);
17 }
18
19 exists(0:r1=1 /\ 0:r2=1 /\ 1:r3=1)
\end{verbatim}
\caption{Causality Test Case 2}
\label{lst:Causality Test Case 2}
\end{listing}

\subsubsection{Causality Test Case 3}
\label{app:Causality Test Case 3}

Listing~\ref{lst:Causality Test Case 3}
shows causality test case 3, for which the \co{r1 == r2 == r3 == 1} result
is to be allowed.
The herd7 tool finds this OOTA cycle.

\begin{listing}[tbp]
\begin{verbatim}
 1 C oota-causality-3
 2 {
 3   [x] = 0;
 4   [y] = 0;
 5 }
 6
 7 P0(volatile atomic_int *x, volatile atomic_int *y) {
 8   int r1 = atomic_load_explicit(x, memory_order_relaxed);
 9   int r2 = atomic_load_explicit(x, memory_order_relaxed);
10   if (r1 == r2)
11     atomic_store_explicit(y, 1, memory_order_relaxed);
12 }
13
14 P1(volatile atomic_int *x, volatile atomic_int *y) {
15   int r3 = atomic_load_explicit(y, memory_order_relaxed);
16   atomic_store_explicit(x, r3, memory_order_relaxed);
17 }
18
19 P2(volatile atomic_int *x, volatile atomic_int *y) {
20   atomic_store_explicit(x, 2, memory_order_relaxed);
21 }
22
23 exists(0:r1=1 /\ 0:r2=1 /\ 1:r3=1)
\end{verbatim}
\caption{Causality Test Case 3}
\label{lst:Causality Test Case 3}
\end{listing}

\subsubsection{Causality Test Case 4}
\label{app:Causality Test Case 4}

Listing~\ref{lst:Causality Test Case 4}
shows causality test case 4, for which the \co{r1 == r2 == 1} result
is to be forbidden.
The herd7 tool finds this OOTA cycle, but flags it as such using \co{S08}.

\begin{listing}[tbp]
\begin{verbatim}
 1 C oota-causality-4
 2 {
 3   [x] = 0;
 4   [y] = 0;
 5 }
 6
 7 P0(atomic_int *x, atomic_int *y) {
 8   int r1 = atomic_load_explicit(x, memory_order_relaxed);
 9   atomic_store_explicit(y, r1, memory_order_relaxed);
10 }
11
12 P1(atomic_int *x, atomic_int *y) {
13   int r2 = atomic_load_explicit(y, memory_order_relaxed);
14   atomic_store_explicit(x, r2, memory_order_relaxed);
15 }
16
17 exists(0:r1=1 /\ 1:r2=1)
\end{verbatim}
\caption{Causality Test Case 4}
\label{lst:Causality Test Case 4}
\end{listing}

\subsubsection{Causality Test Case 5}
\label{app:Causality Test Case 5}

Listing~\ref{lst:Causality Test Case 5}
shows causality test case 5, for which the \co{r1 == r2 == 1 && r3 == 0}
result is to be forbidden.
The herd7 tool finds this OOTA cycle, but flags it as such using \co{S08}.

\begin{listing}[tbp]
\begin{verbatim}
 1 C oota-causality-5
 2 {
 3   [x] = 0;
 4   [y] = 0;
 5   [z] = 0;
 6 }
 7
 8 P0(atomic_int *x, atomic_int *y, atomic_int *z) {
 9   int r1 = atomic_load_explicit(x, memory_order_relaxed);
10   atomic_store_explicit(y, r1, memory_order_relaxed);
11 }
12
13 P1(atomic_int *x, atomic_int *y, atomic_int *z) {
14   int r2 = atomic_load_explicit(y, memory_order_relaxed);
15   atomic_store_explicit(x, r2, memory_order_relaxed);
16 }
17
18 P2(atomic_int *x, atomic_int *y, atomic_int *z) {
19   atomic_store_explicit(z, 1, memory_order_relaxed);
20 }
21
22 P3(atomic_int *x, atomic_int *y, atomic_int *z) {
23   int r3 = atomic_load_explicit(z, memory_order_relaxed);
24   atomic_store_explicit(x, r3, memory_order_relaxed);
25 }
26
27 exists(0:r1=1 /\ 1:r2=1 /\ 3:r3=0)
\end{verbatim}
\caption{Causality Test Case 5}
\label{lst:Causality Test Case 5}
\end{listing}

\subsection{Additional Litmus Tests}
\label{app:Additional Litmus Tests}

\subsubsection{Non-LB Litmus Tests}
\label{app:Non-LB Litmus Tests}

Listing~\ref{lst:Non-LB Litmus Test}
shows a litmus test that can produce an OOTA cycle, but that is not
a member of the load-buffering (LB) class of litmus tests.
Suppose that lines~9 and~10 both somehow return the value 21.
Then line~12 will store the value 42 to \co{x}, which will cause
line~17 to store 21 to \co{y} and line~22 to store 21 to \co{z},
thus justifying the values initially loaded by lines~9 and~10.

\begin{listing}[tbp]
\begin{verbatim}
 1 C oota-non-lb
 2 {
 3   [x] = 2;
 4   [y] = 1;
 5   [z] = 1;
 6 }
 7
 8 P0(atomic_int *x, atomic_int *y, atomic_int *z) {
 9   int r0 = atomic_load_explicit(y, memory_order_relaxed);
10   int r1 = atomic_load_explicit(z, memory_order_relaxed);
11   int r2 = r0 + r1;
12   atomic_store_explicit(x, r2, memory_order_relaxed);
13 }
14
15 P1(atomic_int *x, atomic_int *y, atomic_int *z) {
16   int r3 = atomic_load_explicit(x, memory_order_relaxed);
17   atomic_store_explicit(y, r3 / 2, memory_order_relaxed);
18 }
19
20 P2(atomic_int *x, atomic_int *y, atomic_int *z) {
21   int r4 = atomic_load_explicit(x, memory_order_relaxed);
22   atomic_store_explicit(z, r4 / 2, memory_order_relaxed);
23 }
24
25 locations [0:r0;0:r1;0:r2;1:r3;2:r4]
26 exists(x=42 /\ y=21 /\ z=21)
\end{verbatim}
\caption{Non-LB Litmus Test}
\label{lst:Non-LB Litmus Test}
\end{listing}
